---
title: "Weather Events: Economic & Health Risks"
author: "Scott K"
date: "8/3/2020"
output: html_document
---

### Synopsis

In this report, we aim to assess the destruction caused by weather events
between 1950 and 2011.  Our goal is to determine which event causes the most 
damage, as measured by both health risks (injures and deaths) and financial
risks (property and crop damage).  From these data, we have determined that,
across of the documented events that Tornadoes are most harmful with respect to 
both health and economic consequences.

### Loading and Processing the Raw Data

#### Reading in the data

We loaded in the necessary libraries and read in the data from the included
zip archive.  The data is in .csv format and the specific variables that we 
need do not need any tidying

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(ggpubr)

data_file <- "repdata_data_StormData.csv.bz2"

data <- read.csv(data_file)
```

After reading in the data, we check the first few rows of the dataset.

```{r}
dim(data)

head(data)
```

#### Processing the data

We first want to answer the question, which types of events are most harmful with 
respect to population health?  To assess this, we must summarize the variables, 
injuries and deaths, by weather event.

```{r}
data_deaths <- data %>%
                group_by(event = EVTYPE) %>%
                summarize(deaths = sum(FATALITIES)) %>%
                arrange(desc(deaths)) %>%
                top_n(10, wt = deaths)

data_injuries <- data %>%
        group_by(event = EVTYPE) %>%
        summarize(injuries = sum(INJURIES)) %>%
        arrange(desc(injuries)) %>%
        top_n(10, wt = injuries)
```

Next, we want to answer the question, which type of events have the greatest
economic consequences?  To assess this, we must summarize the variables, crop 
damange and property damage.

```{r}
data_damage <- data %>%
        group_by(event = EVTYPE) %>%
        summarize(property_damage = sum(PROPDMG)/1000,
                crop_damage = sum(CROPDMG)/1000,
                total_damage = (property_damage + crop_damage)) %>%
        arrange(desc(total_damage)) %>%
        top_n(10, wt = total_damage)

```

### Results

#### Events most harmful to population health

In order to show the events with the greatest health risks, we will list the 
top 10 in terms of both deaths and injuries.

```{r}
data_deaths

data_injuries
```

We will also produce a barchart showing that Torndaoes, clearly, cause the 
greatest risk to the population's health.

```{r}
death_plot <- ggplot(data_deaths, aes(event, deaths)) +
        geom_bar(stat = "identity", aes(fill = event)) +
        ylab("Total Deaths") +
        xlab("Event Type") +
        theme(axis.text.x = element_text(angle = 90, size = 6),
                legend.position = "none",
                axis.title.x = element_text(color = "blue", 
                        size = 12),
                axis.title.y = element_text(color = "blue", 
                        size = 12))
        
injuries_plot <- ggplot(data_injuries, aes(event, injuries)) +
        geom_bar(stat = "identity", aes(fill = event)) +
        ylab("Total Injuries") +
        xlab("Event Type") +
        theme(axis.text.x = element_text(angle = 90, size =6),
                legend.position = "none",
        axis.title.x = element_text(color = "blue", 
                size = 12),
        axis.title.y = element_text(color = "blue", 
                size = 12))

ggarrange(death_plot, injuries_plot, ncol = 2, 
        labels = c("Total Deaths per Event Type", 
                "Total Injuries per Event Type"),
        font.label = list(color = "blue"))
```

#### Events with gratest economic consequences

We will do the same for economic risk.  Here is the top 10:

```{r}
data_damage
```

And here is the barchart showing, again, that Tornadoes cause the most risk with 
regards to finances, as well.

```{r}

ggplot(data_damage, aes(event, total_damage)) +
        geom_bar(stat = "identity", aes(fill = event)) +
        ylab("Total Damage (,000s)") +
        xlab("Event Type") +
        ggtitle("Total Damage per Event Type") +
        theme(axis.text.x = element_text(angle = 90, size =6),
                legend.position = "none", 
                plot.title = element_text(hjust = 0.5, color = "blue", 
                        size = 16, face = "bold"),
                axis.title.x = element_text(color = "blue", 
                        size = 12, face ="bold"),
                axis.title.y = element_text(color = "blue", 
                        size = 12, face ="bold"))
```



